Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
Simplify list of key ideas in S1

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{is designed to accept the lazy} semantics of
linearity. \ROUNDTWO{Our system achieves this via a novel combination
  of features: explicit tracking linear dependencies of variables
  through \emph{usage environments}
  (Section~\ref{sec:usage-environments}), codifying that lazy variable
  bindings do not consume resources upon definition, but rather when
  the variables are consumed;
  a distinct treatment of case
  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
  identifying that linear resources captured by case expressions in
  WHNF can be safely used in branches since the case will not further
  evaluate its scrutinee; and a notion of irrelevant and tagged resources
  (Section~\ref{sec:irrev}), which allow for fine-grained control over
  variable usage. Namely, linear resources occurring in non-WHNF case
  scrutinees can no longer be used in the alternative, while the
  pattern variables (or the case binder) must necessarily be used. } 
%
To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

\noindent (1) Encodes the idea that lazy bindings do not consume resources upon
definition, but rather when the bound variables themselves are consumed.
Consuming a variable with some usage environment $\Delta$ equates to consuming
the variables contained in $\Delta$.
%
(2) Case expressions in WHNF may capture ambient linear resources, but these
resources can be safely used in branches since the case will not further
evaluate its scrutinee. On the other hand, scrutinizing a non-WHNF expression
must be treated more conservatively.
%
(3) Irrelevant resources make existing bindings unusable while forcing the use
of others. Namely, linear resources occurring in non-WHNF case scrutinees can
no longer be used in the alternatives, while the pattern variables or the case
binder must necessarily be used. Tagged resources follow the observation that
some variables must be used jointly or not at all.
}
%
% Linear Core's first key idea is to delay consuming
% resources until a computation that depends on those resources is effectively
% evaluated or returned. Delayed consumption is tracked by annotating relevant
% binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
%
% The second key idea is to have two distinct rules for case
% expressions, \ROUNDTWO{discriminating} on whether the scrutinee is in Weak Head Normal
% Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
% longer available but have yet to be fully consumed
% (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
% resources amongst pattern-bound variables and %^as usage environments
% encode that all linear components of a pattern must be used in tandem.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context and
\ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.

>>>>

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{is designed to accept the lazy} semantics of
linearity. \ROUNDTWO{Our system achieves this via a novel combination
  of features: explicit tracking linear dependencies of variables
  through \emph{usage environments}
  (Section~\ref{sec:usage-environments});
  a distinct treatment of case
  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf});
  and a notion of irrelevant and tagged resources
  (Section~\ref{sec:irrev}), which allow for fine-grained control over
  variable usage.} 
%
To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

\noindent (1) Encodes the idea that lazy bindings do not consume resources upon
definition, but rather when the bound variables themselves are consumed.
Consuming a variable with some usage environment $\Delta$ equates to consuming
the variables contained in $\Delta$.
%
(2) Case expressions in WHNF may capture ambient linear resources, but these
resources can be safely used in branches since the case will not further
evaluate its scrutinee. On the other hand, scrutinizing a non-WHNF expression
must be treated more conservatively.
%
(3) Irrelevant resources make existing bindings unusable while forcing the use
of others. Namely, linear resources occurring in non-WHNF case scrutinees can
no longer be used in the alternatives, while the pattern variables (or the case
binder) must necessarily be used. Tagged resources follow the observation that
some variables must be used jointly or not at all.
}
%
% Linear Core's first key idea is to delay consuming
% resources until a computation that depends on those resources is effectively
% evaluated or returned. Delayed consumption is tracked by annotating relevant
% binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
%
% The second key idea is to have two distinct rules for case
% expressions, \ROUNDTWO{discriminating} on whether the scrutinee is in Weak Head Normal
% Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
% longer available but have yet to be fully consumed
% (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
% resources amongst pattern-bound variables and %^as usage environments
% encode that all linear components of a pattern must be used in tandem.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
the scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context and
\ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$}.
